---
title: "Exercise 6: State Space"
author: "Nik Bates-Haus"
output:
  pdf_document: default
  html_document: default
---

# Activity 6 - State-space models

This activity will explore the state-space framework for modeling
time-series and spatial data sets. Chapter 8 provides a more in-depth
description of the state-space model, but in a nutshell it is based on
separating the process model, which describes how the system evolves in
time or space, from the observation error model. Furthermore, the
state-space model gets its name because the model estimates that true
value of the underlying **latent** state variables.

For this activity we will write all the code, process all the data, and
visualize all the outputs in R, but the core of the Bayesian computation
will be handled by JAGS (Just Another Gibbs Sampler,
<http://mcmc-jags.sourceforge.net>). Therefore, before we get started
you will want to download both the JAGS software and the rjags library,
which allows R to call JAGS. We're also going to install our
`ecoforecastR` package, which has some helper functions we will use.

```{r setup, message = FALSE, warning = FALSE}
library(rjags)
library(daymetr)
#devtools::install_github("EcoForecast/ecoforecastR",force=TRUE)
library(ecoforecastR)

library(tidyverse) # Nik likes tidyverse
```

Next we'll want to grab the data we want to analyze. For this example
we'll use the Google Flu Trends data for the state of Massachusetts,
which we saw how to pull directly off the web in Activity 3.

```{r}
gflu = read.csv("data/gflu_data.txt",skip=11)
time = as.Date(gflu$Date)
y = gflu$Massachusetts
plot(time,y,type='l',ylab="Flu Index",lwd=2,log='y')
```

Next we'll want to define the JAGS code, which we'll do by writing the
code as a string in R. The code itself has three components, the data
model, the process model, and the priors. The **data model** relates the
observed data, y, at any time point to the latent variable, x. For this
example we'll assume that the observation model just consists of
Gaussian observation error.

$$Y_{t} \sim N(X_{t},\tau_{obs})$$

The **process model** relates the state of the system at one point in
time to the state one time step ahead. In this case we'll start with the
simplest possible process model, a random walk, which just consists of
Gaussian process error centered around the current value of the system.

$$X_{t+1} \sim N(X_{t},\tau_{add})$$

Finally, for the priors we need to define **priors** for the initial
condition, the process error, and the observation error.

```{r}
RandomWalk = "
model{
  
  #### Data Model
  for(t in 1:n){
    y[t] ~ dnorm(x[t],tau_obs)
  }
  
  #### Process Model
  for(t in 2:n){
    x[t]~dnorm(x[t-1],tau_add)
  }
  
  #### Priors
  x[1] ~ dnorm(x_ic,tau_ic)
  tau_obs ~ dgamma(a_obs,r_obs)
  tau_add ~ dgamma(a_add,r_add)
}
"
```

Next we need to define the data and priors as a list. For this analysis
we'll work with the log of the Google flu index since the zero-bound on
the index and the magnitudes of the changes appear much closer to a
log-normal distribution than to a normal.

```{r}
data <- list(y=log(y),n=length(y),      ## data
             x_ic=log(1000),tau_ic=100, ## initial condition prior
             a_obs=1,r_obs=1,           ## obs error prior
             a_add=1,r_add=1            ## process error prior
             )
```

Next we need to definite the initial state of the model's parameters for
each chain in the MCMC. The overall initialization is stored as a list
the same length as the number of chains, where each chain is passed a
list of the initial values for each parameter. Unlike the definition of
the priors, which had to be done independent of the data, the
initialization of the MCMC is allowed (and even encouraged) to use the
data. However, each chain should be started from different initial
conditions. We handle this below by basing the initial conditions for
each chain off of a different random sample of the original data.

```{r}
nchain = 3
init <- list()
for(i in 1:nchain){
  y.samp = sample(y,length(y),replace=TRUE)
  init[[i]] <- list(tau_add=1/var(diff(log(y.samp))),  ## initial guess on process precision
                    tau_obs=5/var(log(y.samp)))        ## initial guess on obs precision
}
```

Now that we've defined the model, the data, and the initialization, we
need to send all this info to JAGS, which will return the JAGS model
object.

```{r}
j.model   <- jags.model (file = textConnection(RandomWalk),
                             data = data,
                             inits = init,
                             n.chains = 3)
```

Next, given the defined JAGS model, we'll want to take a few samples
from the MCMC chain and assess when the model has converged. To take
samples from the MCMC object we'll need to tell JAGS what variables to
track and how many samples to take.

```{r, fig.asp = 1.0}
## burn-in
jags.out   <- coda.samples (model = j.model,
                            variable.names = c("tau_add","tau_obs"),
                                n.iter = 1000)
plot(jags.out)
dic.samples(j.model, 2000)
```

Here we see that the model converges rapidly. Since rjags returns the
samples as a CODA object, we can use any of the diagnostics in the R
*coda* library to test for convergence, summarize the output, or
visualize the chains.

Now that the model has converged we'll want to take a much larger sample
from the MCMC and include the full vector of X's in the output

```{r}
jags.out   <- coda.samples (model = j.model,
                            variable.names = c("x","tau_add","tau_obs"),
                                n.iter = 10000)
```

Given the full joint posterior samples, we're next going to visualize
the output by just looking at the **95% credible interval of the
time-series of X's** and compare that to the observed Y's. To do so
we'll convert the coda output into a matrix and then calculate the
quantiles. Looking at colnames(out) will show you that the first two
columns are `tau_add` and `tau_obs`, so we calculate the CI starting
from the 3rd column. We also transform the samples back from the log
domain to the linear domain.

```{r}
time.rng = c(1,length(time))       ## adjust to zoom in and out
out <- as.matrix(jags.out)         ## convert from coda to matrix  
x.cols <- grep("^x",colnames(out)) ## grab all columns that start with the letter x
ci <- apply(exp(out[,x.cols]),2,quantile,c(0.025,0.5,0.975)) ## model was fit on log scale

plot(time,ci[2,],type='n',ylim=range(y,na.rm=TRUE),ylab="Flu Index",log='y',xlim=time[time.rng])
## adjust x-axis label to be monthly if zoomed
if(diff(time.rng) < 100){ 
  axis.Date(1, at=seq(time[time.rng[1]],time[time.rng[2]],by='month'), format = "%Y-%m")
}
ecoforecastR::ciEnvelope(time,ci[1,],ci[3,],col=ecoforecastR::col.alpha("lightBlue",0.75))
points(time,y,pch="+",cex=0.5)
```

Next, lets look at the posterior distributions for `tau_add` and
`tau_obs`, which we'll convert from precisions back into **standard
deviations**.

```{r}
hist(1/sqrt(out[,1]),main=colnames(out)[1])
hist(1/sqrt(out[,2]),main=colnames(out)[2])
```

We'll also want to look at the joint distribution of the two parameters
to check whether the two parameters strongly covary.

```{r, fig.asp = 1.0}
plot(out[,1],out[,2],pch=".",xlab=colnames(out)[1],ylab=colnames(out)[2])
cor(out[,1:2])
```

## Question 1 [A]:

To explore the ability of state space models to generate forecasts (or
in this case, a hindcast) make a copy of the data and **remove the last
40 observations (convert to NA)** and refit the model.

```{r}
set.seed(20260306)
y_hindcast <- c(y[1:(length(y)-40)], rep(NA, 40))
data_hindcast <- list(
  y=log(y_hindcast),
  n=length(y_hindcast),
  x_ic=log(1000),
  tau_ic=100,
  a_obs=1,
  r_obs=1,
  a_add=1,
  r_add=1
)
init_hindcast <- lapply(1:nchain, \(i) {
  y.samp = sample(y_hindcast, length(y_hindcast), replace=TRUE)
  return(list(
      tau_add=1/var(diff(log(y.samp))),
      tau_obs=5/var(log(y.samp))
  ))
})
model_hindcast <- jags.model(
  file = textConnection(RandomWalk),
  data = data_hindcast,
  inits = init_hindcast,
  n.chains = nchain
)
out_hindcast_burnin <- coda.samples(
  model = model_hindcast,
  variable.names = c("tau_add","tau_obs"),
  n.iter = 1500
)
for (var in varnames(out_hindcast_burnin)) {
  gelman.plot(
    out_hindcast_burnin[, var, drop = FALSE]
  )
  abline(h = 1.05, col = "blue", lty = 3)
}
```

> We get to burn-in around iteration 1250 or so.

```{r}
out_hindcast <- coda.samples(
  model = model_hindcast,
  variable.names = c("x","tau_add","tau_obs"),
  n.iter = 10000
)
```

-   Generate a time-series plot for the CI of x that includes all the
    original observed data (as above but zoom the plot on the last \~80
    observations). Use a different color and symbol to differentiate
    observations that were included in the model versus those that were
    converted to NA's.

```{r}
time.rng = c(length(time) - 80, length(time))
out_hindcast_matrix <- as.matrix(out_hindcast)
ci_hindcast <- apply(
  exp(out_hindcast_matrix[, x.cols]),
  2,
  quantile,
  c(0.025, 0.5, 0.975)
)

plot(
  time,
  ci_hindcast[2, ],
  type='n',
  ylim=range(y_hindcast, na.rm=TRUE),
  ylab="Flu Index",
  log='y',
  xlim=time[time.rng]
)

## adjust x-axis label to be monthly if zoomed
if (diff(time.rng) < 100) {
  axis.Date(
    1,
    at=seq(time[time.rng[1]], time[time.rng[2]], by='month'),
    format = "%Y-%m"
  )
}
ecoforecastR::ciEnvelope(
  time,
  ci_hindcast[1,],
  ci_hindcast[3,],
  col=ecoforecastR::col.alpha("lightBlue",0.75)
)
points(time, y_hindcast, pch="+", cex=0.5)
points(
  time[(length(y) - 39):length(y)],
  y[(length(y) - 39):length(y)],
  pch="x",
  cex=0.5,
  col="red"
)
legend(
  "topleft",
  legend = c("Observed", "Hidden"),
  col = c("black", "red"),
  pch = c("+", "x")
)
```

-   Comment on how well the random walk model performed (both accuracy
    and precision) and how it might be modified to improve both these
    criteria.

> Almost all of the hidden points are in the confidence interval, but
> the confidence interval gets very broad very quickly once it can no
> longer see the data. The mean of the model is also flat - it does not
> follow the data at all. So, once the model can no longer see the data,
> it is both inaccurate and imprecise.
>
> The model could be improved in a few ways:
>
> 1.  Lag - temporal PCA on the original time series should give some
>     good ideas for what lags (temporal autocorrelation) are relevant.
>     The likely candidates have to do with incubation time and
>     contagion time for the flu, and perhaps annual.
> 2.  Time - Flu is seasonal so adding a seasonal effect would probably
>     help a lot, as would information about times that people gather
>     together. Furthermore, each year has different flu strains that
>     have different characteristics, so it would help to have a
>     per-year effect for the strains active that year.
>
> So I'd propose a model that incorporates:
>
> -   A time-of-year risk factor, which is just the number of cases (or
>     fraction of population) that had the flu on that date in prior
>     seasons. Oddly, this wants to follow the civil calendar so it can
>     capture things like holidays. This should increase accuracy a lot.
>
> -   Incubation + Contagion Lag - flu takes \~2 days to incubate and is
>     contagious for a week after, so in addition to the existing 1 week
>     lag I'd add a 2 week lag. This would also help the model track
>     rising and falling trends. This should increase precision a lot.
>
> -   Strain - add a per-year hierarchical parameter for how contagious
>     that year's strains are. This should increase accuracy and
>     precision.

## Question 2 [C: Extra Credit]

To look at how observation frequency affects data assimilation, convert
3 out of every 4 observations to NA (i.e. treat the data as
approximately monthly) and refit the model.

```{r}
set.seed(20260306)
y_monthly <- rep(NA, length(y))
y_monthly[seq(1, length(y), 4)] <- y[seq(1, length(y), 4)]
data_monthly <- list(
  y=log(y_monthly),
  n=length(y_monthly),
  x_ic=log(1000),
  tau_ic=100,
  a_obs=1,
  r_obs=1,
  a_add=1,
  r_add=1
)
init_monthly <- lapply(1:nchain, \(i) {
  y_valid <- na.omit(y_monthly)
  y.samp = sample(na.omit(y_valid), length(y_valid), replace=TRUE)
  return(list(
      tau_add=1/var(diff(log(y.samp))),
      tau_obs=5/var(log(y.samp))
  ))
})
model_monthly <- jags.model(
  file = textConnection(RandomWalk),
  data = data_monthly,
  inits = init_monthly,
  n.chains = nchain
)
out_monthly_burnin <- coda.samples(
  model = model_monthly,
  variable.names = c("tau_add","tau_obs"),
  # Note that it takes much longer to burn-in!
  n.iter = 6000
)
for (var in varnames(out_monthly_burnin)) {
  gelman.plot(
    out_monthly_burnin[, var, drop = FALSE] |>
      window(start = 3000)
  )
  abline(h = 1.05, col = "blue", lty = 3)
}
```

> We get to burn-in around iteration 4000 or so.

```{r}
out_monthly <- coda.samples(
  model = model_monthly,
  variable.names = c("x","tau_add","tau_obs"),
  n.iter = 10000
)
```

-   Generate a time-series plot for the CI of x that includes the
    observations (as above). Use a different color and symbol to
    differentiate observations that were included in the model versus
    those that were converted to NA's.

```{r}
time.rng = c(length(time) - 104, length(time))
out_monthly_matrix <- as.matrix(out_monthly)
monthly_ci <- apply(
  exp(out_monthly_matrix[, x.cols]),
  2,
  quantile,
  c(0.025, 0.5, 0.975)
)

plot(
  time,
  monthly_ci[2, ],
  type='n',
  ylim=range(y, na.rm=TRUE),
  ylab="Flu Index",
  log='y',
  xlim=time[time.rng]
)

## adjust x-axis label to be monthly if zoomed
if (diff(time.rng) < 100) {
  axis.Date(
    1,
    at=seq(time[time.rng[1]], time[time.rng[2]], by='month'),
    format = "%Y-%m"
  )
}
ecoforecastR::ciEnvelope(
  time,
  monthly_ci[1,],
  monthly_ci[3,],
  col=ecoforecastR::col.alpha("lightBlue",0.75)
)
points(time, y_monthly, pch="+", cex=0.5)
y_unmonthly <- y
y_unmonthly[seq(1, length(y), 4)] <- NA
points(
  time,
  y_unmonthly,
  pch="x",
  cex=0.5,
  col="red"
)
legend(
  "topleft",
  legend = c("Observed", "Hidden"),
  col = c("black", "red"),
  pch = c("+", "x")
)
```

-   Compare the CI between the two runs.

In the hindcast run, the CI was about 300 wide until the model stopped
seeing the data, then ballooned outward.

in the monthly run, the CI fluctuates between about 200 and 400; it
balloons a bit at the end, but nowhere near as much as the hindcast CI.

-   Generate a predicted (median) vs observed plot for the data points
    that were removed

```{r}
predicted <- monthly_ci[2, ]
predicted[seq(1, length(y), 4)] <- NA
observed <- y
observed[seq(1, length(y), 4)] <- NA
plot(predicted, observed, log = "xy")
abline(a=0,b=1,lty=2,col="blue")
```

-   Comment on the accuracy and precision of the state estimates.

> Plotting is nice, but I like numbers:

```{r}
quantile(
  abs(predicted - observed) / observed,
  c(0.025, 0.25, 0.5, 0.75, 0.975),
  na.rm = TRUE
)
```

> So median accuracy of prediction is within 11.4% of observed, 75% is
> within 20% of observed, 97.5% within 56%.
>
> What's the size of the confidence interval?

```{r}
quantile(
  0.5 * (monthly_ci[3, ] - monthly_ci[1, ]) / observed,
  c(0.025, 0.25, 0.5, 0.75, 0.975),
  na.rm = TRUE
)
```

> So median precision is ±70%, which is pretty low.

-   How does the reduction in data volume affect the parameter estimates
    (taus)

> Here is the distribution of tau_add and tau_obs for the full data:

```{r}
data.frame(
  tau_add = quantile(
    out[, "tau_add"],
    c(0.025, 0.25, 0.5, 0.75, 0.975)
  ),
  tau_obs = quantile(
    out[, "tau_obs"],
    c(0.025, 0.25, 0.5, 0.75, 0.975)
  )
)
```

> And tau_add / tau_obs for the monthly data:

```{r}
data.frame(
  tau_add = quantile(
    out_monthly_matrix[, "tau_add"],
    c(0.025, 0.25, 0.5, 0.75, 0.975)
  ),
  tau_obs = quantile(
    out_monthly_matrix[, "tau_obs"],
    c(0.025, 0.25, 0.5, 0.75, 0.975)
  )
)
```

So the change from weekly to monthly brought tau_add (process model
precision) from 23 ± 3 to 13 ± 3, and tau_obs (data model precision)
from 62 ± 11 to 10 ± 4.5. Precision is $1/\sigma^2$, so this reduction
in precision is an *increase* in variance, which matches expectations:
decreasing data increases variance and decreases precision.

# Dynamic Linear Models

The random walk model can easily be generalized to more sophisiticated
models describing the dynamics of the system. One simple but useful
extension is the class of dynamic linear models (DLMs) -- linear models
where the future state depends on the current state and other
covariates, $z_t$

$$X_{t+1} \sim N(X_t + \beta_0 + \beta_1 z_t + \beta_{X} X_{t}, \tau_{add})$$

where $\beta_0$ is the intercept, $\beta_1$ is the slope of the
covariate effect, and $\beta_{X}$ is the slope of the initial condition
effect, expressed as a deviation from the random walk default (i.e. the
actual slope is $1 + \beta_X$). Rather than implement this model in JAGS
directly, we're going to rely on the `ecoforecastR` package, which
accepts a `lm` like syntax for specifying covariates (with the notable
exception that the response variable, which is our latent X, is not
specified explicitly). Here we're going to use the Daymet product to get
daily weather estimates, and then use daily minimum temperature (Tmin)
as the covariate in our influenza model

```{r}
## grab weather data
df <- daymetr::download_daymet(site = "Boston",
                lat = 42.36,
                lon = -71.06,
                start = 2003,
                end = 2016,
                internal = TRUE)$data
df$date <- as.Date(paste(df$year,df$yday,sep = "-"),"%Y-%j")
data$Tmin = df$tmin..deg.c.[match(time,df$date)]
Tbar = mean(data$Tmin,na.rm=TRUE)
data$Tmin = df$tmin..deg.c.[match(time,df$date)] - Tbar ## covert Tmin to anomalies

## fit the model
ef.out <- ecoforecastR::fit_dlm(model=list(obs="y",fixed="~ 1 + X + Tmin"),data)
names(ef.out)
```

```{r}
plot(df$date, df$tmin, pch=".", col="blue")
```

```{r}
for (var in varnames(ef.out$params)) {
  gelman.plot(
    ef.out$params[, var, drop = FALSE]
  )
  abline(h = 1.05, col = "blue", lty = 3)
}
```

The package returns a list with four elements. `params` and `predict`
are both the same mcmc.list objects we get back from JAGS, only split
between the parameters and the latent state variables, respectively, to
make it easier to perform diagnostics and visualizations:

```{r, fig.asp = 1.0}
## parameter diagnostics
params <- window(ef.out$params,start=1000) ## remove burn-in
plot(params)
summary(params)
cor(as.matrix(params))
pairs(as.matrix(params))

## confidence interval
out <- as.matrix(ef.out$predict)
ci <- apply(exp(out),2,quantile,c(0.025,0.5,0.975))
plot(time,ci[2,],type='n',ylim=range(y,na.rm=TRUE),ylab="Flu Index",log='y',xlim=time[time.rng])
## adjust x-axis label to be monthly if zoomed
if(diff(time.rng) < 100){ 
  axis.Date(1, at=seq(time[time.rng[1]],time[time.rng[2]],by='month'), format = "%Y-%m")
}
ecoforecastR::ciEnvelope(time,ci[1,],ci[3,],col=ecoforecastR::col.alpha("lightBlue",0.75))
points(time,y,pch="+",cex=0.5)
```

The JAGS model that was fit 'under the hood' is returned as `model`
which we can view as:

```{r, echo=FALSE}
strsplit(ef.out$model,"\n",fixed = TRUE)[[1]]
```

This code illustrates a few things:

-   The "Priors" section is identical to our earlier random walk model
-   The "Random Effects" section, which is currently commented out,
    illustrates that the `ecoforcastR::fit_dlm` function supports random
    effects, which can be turned on via the `model$random` argument
-   The "Fixed Effects" section contains additional priors for our fixed
    effects as well as priors on the means (mu) and precisions (tau) of
    the covariates.
-   The "Data Model" section is the same as in our random walk except
    for the addition of code for the means of the covariates. This code
    is here as a very simple missing data model -- any time the
    covariate is observed it is used to estimate the mean and precision,
    but any time the covariate is missing (NA) it is imputed.
-   The "Process Model" is very similar to the random walk, except now
    the expected value (mu) is calculated according to the linear model
    described earlier

Finally, the returned object also includes the `data` that was used to
fit the model.

## Question 3 [A]:

-   Compare the process and observation error estimates and model CI
    between this fit and the original random walk model. How much has
    the residual variance been reduced by?

> Observation error (precision) is in tau_obs
>
> Process error (precision) is in tau_add
>
> Parameters are in ef.out\$params

```{r}
data.frame(
  tau_add = quantile(
    ef.out$params |>
      window(start = 1000) |>
      as.matrix() |>
      _[, "tau_add"],
    c(0.025, 0.25, 0.5, 0.75, 0.975)
  ),
  tau_obs = quantile(
    ef.out$params |>
      window(start = 1000) |>
      as.matrix() |>
      _[, "tau_obs"],
    c(0.025, 0.25, 0.5, 0.75, 0.975)
  )
)
```

> So, tau_add (process precision) is about the same as the model without
> climate data.
>
> But, tau_obs (observation precision) has gone way up. Converting
> precision to variance, the variance in the original, full-data run was
> $1/62=0.016$, the variance with the addition of climate is
> $1/230=0.004$, so ¼ of the original variance, so the original variance
> has been reduced by ¾.

-   Explain and discuss the parameter estimates (betas) from the dynamic
    linear model (what do they mean both biologically and in terms of
    the predictability of the system) and their correlations

> There are three betas estimated in the model: betaIntercept, betaX,
> betaTmin.

```{r}
post_burn_in_params <- ef.out$params |>
      window(start = 1000) |>
      as.matrix()

data.frame(
  betaIntercept = quantile(
    post_burn_in_params |>
      _[, "betaIntercept"],
    c(0.025, 0.25, 0.5, 0.75, 0.975)
  ),
  betaX = quantile(
    post_burn_in_params |>
      _[, "betaX"],
    c(0.025, 0.25, 0.5, 0.75, 0.975)
  ),
  betaTmin = quantile(
    post_burn_in_params |>
      _[, "betaTmin"],
    c(0.025, 0.25, 0.5, 0.75, 0.975)
  )
)
cor(post_burn_in_params[, c("betaIntercept", "betaX", "betaTmin")])
```

> Biologically, betaIntercept is the base flu rate (frequency of
> flu-related search queries, which we presume to be related to the
> frequency of flu-like symptoms, which we presume to be related to the
> frequency of flu). Absent other information, this is what we would
> guess as the flu rate. In terms of the predictability of the system
> betaIntecept is the bar ($\bar{x}$) around which predictions gyre.
>
> Biologically, betaX is how the flu rate at time t tends to differ from
> the flu rate at time t-1 - it is a base rate of growth or decay. In
> terms of the predictability of the system, a positive betaX will tend
> to increase the flu rate, and a negative betaX will tend to decrease
> the flu rate.
>
> Biologically, betaTmin is the influence of daily minimum temperature
> on flu rate. In terms of the predictability of the system, a larger
> magnitude betaTmin means a larger influence of temperature on flu, and
> a negative betaTmin means that lower temperatures correlate with
> higher flu.
>
> The correlations show that betaIntercept and betaX are very highly
> correlated, but with opposite sign (-0.99). Thus, if the mean flu rate
> is high, then flu tends to decline; whereas if mean flu rate is low,
> flu tends to increase. betaTmin is fairly well correlated with the
> other two, also negatively with betaIntercept. Thus, if the mean flu
> rate is low, then colder temperatures will strongly increase flu rate,
> and if the mean flu rate is high, then warmer temperatures will
> strongly decrease flu rate.

-   Because a state-space model returns X's that are close to the Y's,
    metrics such as R2 and RMSE aren't great metrics of model
    performance. Besides looking at the taus, how else could we judge
    which model is doing better (in a way that avoids/penalizes
    overfitting)?

> 1.  We can hide some data from the model and see how it performs on
>     the hidden data. We did that two ways up above: hide the last 40
>     observations, and hide 3 out of 4 observations. Jackknifing and
>     bootstrapping are traditional methods of testing a model's
>     sensitivity to held-out data.

> 2.  Run the model stepwise (outside the JAGS framework) and see how
>     well it does at predicting each timestep.

## Question 4 [B]:

Repeat the process of forecasting the last 40 observations (convert to
NA), this time using the DLM with temperature as a covariate.

-   Generate a time-series plot for the CI that includes the
    observations and both the random walk and DLM models (Hint, think
    about the order you plot in so you can see both models, also
    consider including transpancy [alpha] in the CI color)

```{r}
set.seed(20260306)

data_hindcast2 <- data
data_hindcast2$y <- log(y_hindcast)

out_hindcast2 <- ecoforecastR::fit_dlm(
  model = list(
    obs="y",
    fixed="~ 1 + X + Tmin",
    n.iternumber = 100000 # Raise # of iterations to see if things converge
  ),
  data_hindcast2
)
for (var in varnames(out_hindcast2$params)) {
  gelman.plot(
    out_hindcast2$params[, var, drop = FALSE],
    ylim = range(1, 1.5)
  )
  abline(h = 1.05, col = "blue", lty = 3)
}
```

So it looks like burn-in actually happens around iteration
35,000; I'll use 50,000 to be safe.

```{r}
params2 <- window(out_hindcast2$params, start=50000) # longer burn-in
plot(params2)
summary(params2)
cor(as.matrix(params2))
pairs(as.matrix(params2))
```

Let's look at the predictions:

```{r}
## confidence interval
predicted_hindcast2 <- as.matrix(window(out_hindcast2$predict, start=50000))

ci_hindcast2 <- apply(
  exp(predicted_hindcast2),
  2,
  quantile,
  c(0.025,0.5,0.975)
)

values <- tibble(
  date = as_date(time),
  observed = y_hindcast,
  hidden = c(rep(NA, length(y) - 40), y[(length(y) - 39):length(y)])
) |>
  slice_tail(n = 80) |>
  pivot_longer(cols = c(observed, hidden)) |>
  filter(!is.na(value))

ci <- tibble(
    date = as_date(time),
    ci_low = ci_hindcast[1, ],
    ci_high = ci_hindcast[3, ],
    ci_name = "Random Walk"
  ) |>
  slice_tail(n = 80) |>
  union_all(
    tibble(
      date = as_date(time),
      ci_low = ci_hindcast2[1, ],
      ci_high = ci_hindcast2[3, ],
      ci_name = "DLM"
    ) |>
    slice_tail(n = 80)
  )

ggplot(values, aes(date)) +
  geom_ribbon(
    data = ci,
    mapping = aes(ymin = ci_low, ymax = ci_high, fill = ci_name),
    alpha = 0.5
  ) +
  scale_fill_manual(
    name = "Confidence Interval",
    values = c("Random Walk" = "lightblue", "DLM" = "purple")
  ) +
  geom_point(
    data = values,
    mapping = aes(y = value, shape = name, color = name),
    size = 3
  ) +
  scale_color_manual(
    name = "Observations",
    values = c("hidden" = "red", "observed" = "black")
  ) +
  scale_shape_manual(
    name = "Observations",
    values = c("hidden" = "x", "observed" = "+")
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -30, hjust = 0, vjust = 1)) +
  scale_y_continuous(name = "Flu Index", transform = "log10") +
  ggtitle("Predicted vs. Observed Flu Index") +
  xlab("Date")
```

-   Comment on how well the DLM model performed (both accuracy and
    precision) relative to the random walk and the true observations.
    How could the model be further improved?

With the observed data, the accuracy of the two models is the same. Once
the data becomes hidden, the DLM model has more values outside the CI,
so it is less accurate than the random walk.

With the observed data, the precision of the DLM model is better than
the random walk, it looks like by up to 50% better. Once the data is
hidden, the DLM CI grows more slowly and hits a limit within about 3
months, then follows the overall trend of the data. So, overall, the DLM
is much more precise than the Random Walk.

> For improvements, I think the minimum temperature is a good input,
> but I'm sticking with my previous suggestions for improvement:
>
> -   A time-of-year risk factor, which is just the number of cases (or
>     fraction of population) that had the flu on that date in prior
>     seasons. Oddly, this wants to follow the civil calendar so it can
>     capture things like holidays. This should increase accuracy a lot.
>
> -   Incubation + Contagion Lag - flu takes \~2 days to incubate and is
>     contagious for a week after, so in addition to the existing 1 week
>     lag I'd add a 2 week lag. This would also help the model track
>     rising and falling trends. This should increase precision a lot.
>
> -   Strain - add a per-year hierarchical parameter for how contagious
>     that year's strains are. This should increase accuracy and
>     precision.

# Next steps

Apply these modeling approaches to you own time-series data! As a simple
place to start note that you can fit the basic Random Walk model using
`fit_dlm` just by setting `fixed = ""`. Also, as with standard `lm`
syntax, you can suppress the intercept by including -1 in fixed, specify
interaction terms using multiplication (e.g. X \* Tmin), and express
polynomials both on X and on covariates (e.g. X\^2 or Tmin\^2). The
latter allows you to construct models with stabilizing feedbacks, for
example:

$$ N_{t+1} = N_t + rN_t \left( 1 + {{N_t}\over{K}} \right) = (1+r)N_t + {{r}\over{K}}N_t^2$$
can be expressed as fixed = "-1 + X + X\^2" where $\beta_X = r$ and
$\beta_{X^2} = r/K$.

Within the ecoforecastR package, the ParseFixed function (which is used
by fit_dlm) can also construct text strings for process models, priors,
and missing data models that can be inserted into other JAGS models,
which allows you to easily construct non-Gaussian dynamic generalized
linear mixed models (DGLMMs), data fusion models, or more complex
nonlinear models.
