---
title: "Model Assessment"
output:
  pdf_document: default
  html_document: default
---

# Model Assessment

In this activity we will use a series of visualizations and statistical
measures to assess the performance of our Super Simple Ecosystem Model
at the Metolius site.

Let's start by loading the ensemble output from the previous lab and the
observed flux data for the site.

```{r}
## load libraries
library(plotrix)
library(rpart)
library(randomForest)
library(scoringRules)
library(tidyverse)

## load SSEM output
# loads "NEE.ens" "inputs"
load("data/Ex10.output.RData")
# loads "NEE.pf"
load("data/Ex10.output.pf.RData")

## load flux tower data
L4 <- read.csv("data/AMF_USMe2_2005_L4_h_V002.txt", header = TRUE, na.strings = "-9999")
L4[L4 == -9999] <- NA
```

## Model vs. Data

In the following section we will begin with some basic diagnostic plots
and statistics assessing the predicted NEE by our simple ecosystem
model. Specifically, we will calculate the Root Mean Square Error
(RMSE), bias, correlation coefficient, and regression slopes of the
relationship between the observed and predicted NEE for both the
original ensemble and the particle filter. We will also generate scatter
plots of predicted vs. observed values.

```{r}
## Calculate ensemble means & apply QAQC
qaqc <- (L4$qf_NEE_st == 0)
NEE.ens.bar <- -apply(NEE.ens[, ], 1, mean)
NEE.pf.bar <- -apply(NEE.pf[, ], 1, mean)
E <- NEE.ens.bar[qaqc]
P <- NEE.pf.bar[qaqc]
O <- L4$NEE_st_fMDS[qaqc]

## Model vs obs regressions
NEE.ens.fit <- lm(O ~ E)
NEE.pf.fit <- lm(O ~ P)

## performance stats
stats <- as.data.frame(matrix(NA, 4, 2))
rownames(stats) <- c("RMSE", "Bias", "cor", "slope")
colnames(stats) <- c("ens", "pf")
stats["RMSE", "ens"] <- sqrt(mean((E - O)^2))
stats["RMSE", "pf"] <- sqrt(mean((P - O)^2))
stats["Bias", "ens"] <- mean(E - O)
stats["Bias", "pf"] <- mean(P - O)
stats["cor", "ens"] <- cor(E, O)
stats["cor", "pf"] <- cor(P, O)
stats["slope", "ens"] <- coef(NEE.ens.fit)[2]
stats["slope", "pf"] <- coef(NEE.pf.fit)[2]
knitr::kable(stats)

## predicted-observed
plot(E, O, pch = ".", xlab = "ensemble", ylab = "observed", main = "NEE (umol/m2/sec)")
abline(0, 1, col = 2, lwd = 2)
abline(NEE.ens.fit, col = 3, lwd = 3, lty = 2)
legend("bottomright", legend = c("obs", "1:1", "reg"), col = 1:3, lwd = 3)

plot(P, O, pch = ".", xlab = "particle filter", ylab = "observed", main = "NEE (umol/m2/sec)")
abline(0, 1, col = 2, lwd = 2)
abline(NEE.pf.fit, col = 3, lwd = 3, lty = 2)
legend("bottomright", legend = c("obs", "1:1", "reg"), col = 1:3, lwd = 3)
```

### Question 1

Question 1 [A]: Which version of the model performed better? Do the
statistics or plots give any indication about what parameters might need
to be fixed, or processeses refined in the model?

> The particle filter performed slightly better acrosss all metrics. The
> stats / plots don't really show *why* the pf did better, or what to
> tune or change to improve either model.

### Question 2

Question 2 [B]: Plot a time-series of observed and modeled **daily
mean** NEE. Make sure to use the gap-filled NEE estimates, since flux
data are not missing at random. Hint: you can use `tapply` and the day
of year column in the data

```{r}
nee_df <- tibble(
  obs = L4$NEE_st_fMDS,
  ens = NEE.ens.bar,
  pf = NEE.pf.bar,
  qaqc = (L4$qf_NEE_st == 0),
  month = L4$Month,
  day = L4$Day,
  hour = L4$Hour,
  timestamp = as_datetime(paste0(
    "2005-", month, "-", day, " ",
    floor(hour), ":", 60 * hour%%1, ":00"
  ))
) |>
  # Remove observations that fail qaqc
  mutate(obs = if_else(qaqc, obs, NA)) |>
  select(timestamp, obs, ens, pf, qaqc)

nee_daily <- nee_df |>
  mutate(day = date(timestamp)) |>
  group_by(day) |>
  summarize(
    obs = mean(obs, na.rm = TRUE),
    ens = mean(ens),
    pf = mean(pf)
  )

nee_daily |>
  pivot_longer(cols = c(obs, ens, pf)) |>
  ggplot(aes(day, value, color = name)) +
  geom_line(na.rm = TRUE) +
  theme_bw() +
  ylab(bquote("NEE" ~ (umol %.% m^-2 %.% sec^-1))) +
  scale_color_discrete(name = "Legend") +
  ggtitle("Daily Mean NEE")

```

## Taylor diagram

Next, let's use a Taylor diagram to pull our summary statistics together
into one plot. One of the advantages of the Taylor diagram is that it
makes it simpler to visually diagnose the relative differences in model
performance, especially when comparing multiple models or different
versions of the same model. In the figure below we'll begin by plotting
the ensemble, the particle filter, and the climatology. While not
common, the Taylor diagram also provides a way of expressing model and
data uncertainty in the plot by plotting ensemble estimates of both.
Below we add all 200 members of the model ensemble, as well as a Monte
Carlo estimate of observation error in the flux data. The latter is
derived based on the research by Richardson et al (2006), who showed
that eddy covariance data has a non-symmetric heteroskedastic, Laplace
distribution. The non-symmetric part refers to the fact that there is
greater error in positive fluxes (= respiration, typically nocturnal
measurments) than in negative ones.

```{r}
par(pty = "s") # taylor diagrams only make sense if circular

## Taylor diagrams: Ensemble
taylor.diagram(ref = O, model = E, normalize = TRUE, ref.sd = TRUE)

## add full ensemble
for (i in 1:ncol(NEE.pf)) {
  taylor.diagram(ref = O, model = -NEE.ens[qaqc, i], col = 2, pch = ".", add = TRUE, normalize = TRUE)
}

# Taylor Diagram: particle filter
taylor.diagram(ref = O, model = P, add = TRUE, normalize = TRUE, col = 3)

## add data uncertainty
rlaplace <- function(n, mu, b) {
  return(mu + ifelse(rbinom(n, 1, 0.5), 1, -1) * rexp(n, b))
}
beta <- ifelse(O > 0, 0.62 + 0.63 * O, 1.42 - 0.19 * O) # Heteroskedasticity, parameters from Richardson et al 2006
for (i in 1:200) {
  x <- rlaplace(length(O), O, beta)
  taylor.diagram(ref = O, model = x, col = 5, add = TRUE, normalize = TRUE)
}
legend("topright", legend = c("ens", "PF", "obsUncert"), col = 2:5, pch = 20, cex = 0.7)
```

### Question 3

Question 3 [A]: What did you learn about model performance from the
Taylor diagram?

> The ensemble members differ significantly in sd, but are pretty
> tightly clustered in correlation.
> RMSE scales with sd. The aggregate ensemble prediction and the 
> Particle Filter prediction are very close, but far from center.
> Observed uncertainty is small.

### Question 4

Question 4 [B]: How do our simple models compare to the ensemble of
ecosystem models in Figure 7 of Schwalm et al 2010 "A model-data
intercomparison of CO2 exchange across North America: results from the
north american carbon program site synthesis". J. Geophys. Res. ?

> Not so great. Most models in Schwalm (2010) have 0.5 < RMSE < 1.0,
> 0.5 < sdr < 1.0, and 0.85 > corr > 0.5. Our simple models come in
> at ~1.2, ~1.4, and ~0.65, respectively.

## Bayesian p-values

Next, let's look at the patterns in the Bayesian 'p-values', which is
essentially a plot of the quantiles of the observed values relative to
the predictive distributions. The ideal distribution of quantiles is
flat (values show up as frequently as predicted), overcalibrated models
tend to have too many observations near 50%, while poorly calibrated
models will tend to produce a lot of 0's and 1's (consistently under- or
over-predicting).

```{r}
O <- L4$NEE_st_fMDS ## observed
pval.pf <- 0
for (i in 1:nrow(NEE.pf)) {
  pval.pf[i] <- sum(O[i] > -NEE.pf[i, ]) / ncol(NEE.pf) ## quantiles of the particle filter
}
plot(pval.pf) ## quantile 'residuals'
hist(pval.pf, probability = TRUE) ## quantile distribution (should be flat)
```

### Question 5

Question 5 [A]: How do **both** the ensemble and particle filter perform
in terms of the predicted quantiles?

> The particle filter p-values are mostly 1, followed by some that are 0,
> and only a very few that are in between. This indicates that the
> particle filter consistently over-predicts NEE.

```{r}
pval.ens <- 0
for (i in 1:nrow(NEE.ens)) {
  pval.ens[i] <- sum(O[i] > -NEE.ens[i, ]) / ncol(NEE.ens)
}
plot(pval.ens)
hist(pval.ens, probability = TRUE)

```

> The ensemble p-values are mostly 1, followed by some that are 0,
> then a few in the 0.8 - 1.0 range, and only a very few others.
> This indicates that the particle filter consistently over-predicts NEE.

## CRPS

Building on our ideas related to model skill scores (e.g. accuracy vs
precision) and our ideas about assessing probabilistic forecasts (e.g.
Bayesian p-values), is the idea of skill scores for probabalistic
forecasts. A number of such scores exist, with one of the more popular
being the Continuous Rank Probability Score (CRPS).

$$CRPS = {1 \over m} \sum_{i=1}^m|X_i-y| 
- {1 \over {2m^2}}\sum_{i=1}^m\sum_{j=1}^m|X_i-X_j|$$

Where y is the data and the X's are ensemble members within the
prediciton. The first term in CRPS represents the mean absolute error.
The second applies a penalty based on the magnitude of the ensemble
spread. To start, let's apply CRPS to the particle filter forecast,
which returns a score for every data point. To summarize and visualize
this we then calculate the average CRPS, plot CRPS over time, and
calculate a histogram to visualize the distribution.

```{r}
crps <- scoringRules::crps_sample(y = O, dat = NEE.pf)
time <- as.Date(L4$DoY, origin = "2005-01-01")
mean(crps)
plot(time, crps)
hist(crps)
```

It's not particularly easy to see how the model is doing from the
timeseries, so we also calculate the mean diurnal cycle and the
relationship between model skill and flux

```{r}
### diurnal cycle of model skill
crps.diurnal <- tapply(crps, L4$Hour, mean)
time.of.day <- as.numeric(names(crps.diurnal))
plot(time.of.day, crps.diurnal)

## CRPS skill as a function of observed NEE
plot(O, crps)
```

### Question 6

Question 6 [A]: Compare the CRPS for the ensemble and PF models. How do
they perform overall (mean) in terms of their ability to produce the
correct seasonal and diurnal cycles? Explain the pattern of the
relationship between NEE and CRPS.

```{r}
crps.ens <- scoringRules::crps_sample(y = O, dat = NEE.ens)
mean(crps.ens)
plot(time, crps.ens)
hist(crps.ens)
crps.diurnal.ens <- tapply(crps.ens, L4$Hour, mean)
plot(time.of.day, crps.diurnal.ens)
plot(O, crps)
```

> The means are very close (7.46 vs 7.45), but this really doesn't
> tell us much.
>
> The histograms of CRPS scores look very similar between the two,
> with the vast bulk of scores in the 0-5 range, with a tail up to 50.
> This tells us that the distribution or CRPS scores is very similar
> between the two, but not whether those scores are correlated.
>
> Seasonally, for both the ensemble and the particle filter, CRPS is
> lower outside the growing season and higher in the growing season.
>
> With respect to time of day, crps increases dramatically during
> daylight hours. For both models there's a curious dip in CRPS
> around sunrise and sunset; an interesting thing to look into.
>
> For both models, CRPS rises very quickly as observed NEE drops < 0.
> Going back the plot of NEE for question 2, there's a long period
> between June - Oct during which observed NEE declines significantly
> (gets closer to 0), whereas both models still predict that NEE
> will be high (away from 0).
> 
> This indicates that the models are missing an important driver of
> NEE that causes it to drop off during this period.

## Mining the Residuals

In the final section we'll use a few off-the-shelf data mining
approaches to look at the model residuals and ask what parts of our
input space are associated with the largest model error. Note that we
are not limited to just examining the effects of the model inputs, we
might also look at other potential drivers that are not included in our
model, such as soil moisture, to ask if model error is associated with
our failure to include this (or other) drivers. Alternatively, we could
have looked at other factors such as the time of day or even other model
variables (e.g. is model error higher when LAI is larger or small?)

Of the many algorithms out there we'll look at two: the Classification
and Regression Tree (CART) model and the Random Forest model. For both
we'll define our error metric as $(E-O)/beta$, where beta is the
parameter equivalent to the variance in Laplace distribution.
Specifically, we're using the heteroskedastic observation error to
reweight the residuals to account for the fact that large residuals at
times of high flux is likely due to high measurement error. Thus the
errors can be interpreted as similar to the number of of standard
deviations.

The CART model is a classification algorithm which will build a tree
that discretely classifies when the model has high and low error.

The Random Forest model is more like a response surface. The Random
Forest will generate 'partial dependence' plots, which indicate the
importance of each factor across its range, as well as an overall
estimate of the importance of each factor in the model error.

The key thing to remember in all these plots is that we're modelling the
RESIDUALS in order to diagnose errors, not modeling the NEE itself.

```{r, fig.height=6}
## define error metric and dependent variables
O <- L4$NEE_st_fMDS[qaqc]
err <- (E - O) / beta
x <- cbind(inputs$PAR[qaqc], inputs$temp[qaqc])
colnames(x) <- c("PAR", "temp")
smp <- sample.int(length(err), 1000) ## take a sample of the data since some alg. are slow

### Classification tree
rpb <- rpart(err ~ x) ## bias
plot(rpb)
text(rpb)
e2 <- err^2
rpe <- rpart(e2 ~ x) ## sq error
plot(rpe)
text(rpe)

## Random Forest
rfe <- randomForest(x[smp, ], abs(err[smp]))
rfe$importance
partialPlot(rfe, x[smp, ], "PAR")
partialPlot(rfe, x[smp, ], "temp")
```

### Question 7

Question 7 [A]: Overall, which driver is most important in explaining
model error? What conditions are most associated with model success?
With model failure? Where do these results reinforce conclusions we
reached earlier and where do they shine light on new patterns you may
have missed earlier?

> For the decision tree, the top node in both cases is temperature, and
> most splits are based on temperature. For the random forest, splitting
> on temperature also yields greater subtree purity, so here again
> temperature is more important.
>
> Low temperature and low sunlight are most correlated with success ...
> so, the model is best at night and in winter when NEE is at minimum.
> The model is worst when it's bright and hot, which limits its utility.
>
> These restults reinforce that the model is missing an important driver
> in June through August, when it is bright and hot. This driver sends
> observed NEE down, whereas without it the model predicts high NEE.

## Functional Responses

In this section we look at how well the model performed by assessing the
modeled relationships between inputs and outputs and comparing that to
the same relationship in the data. The raw relationships are very noisy,
as many covariates are changing beyond just the single input variable we
are evaluating, so in addition we calculate binned means for both the
model and data.

```{r}
## raw
plot(inputs$temp[qaqc], O, pch = ".", ylab = "NEE")
points(inputs$temp[qaqc], E, pch = ".", col = 2)

## binned
nbin <- 25
# PAR = inputs$PAR[qaqc]
# x = seq(min(PAR),max(PAR),length=nbin)
Tair <- inputs$temp[qaqc]
xd <- seq(min(Tair), max(Tair), length = nbin)
xmid <- xd[-length(xd)] + diff(xd)
bin <- cut(Tair, xd)
Obar <- tapply(O, bin, mean, na.rm = TRUE)
Ose <- tapply(O, bin, std.error, na.rm = TRUE)
Ebar <- tapply(E, bin, mean, na.rm = TRUE)
Ese <- tapply(E, bin, std.error, na.rm = TRUE)
OCI <- -cbind(Obar - 1.96 * Ose, Obar, Obar + 1.96 * Ose)
ECI <- -cbind(Ebar - 1.96 * Ese, Ebar, Ebar + 1.96 * Ese)
rng <- range(rbind(OCI, ECI))

col2 <- ecoforecastR::col.alpha("darkgrey", 0.9)
col1 <- ecoforecastR::col.alpha("lightgrey", 0.6)

plot(xmid, Obar, ylim = rng, type = "n", xlab = "Air Temperature (C)", ylab = "NEP (umol/m2/s)", cex.lab = 1.3)
ecoforecastR::ciEnvelope(xmid, ECI[, 1], ECI[, 3], col = col2)
lines(xmid, ECI[, 2], col = "black", lwd = 4)
ecoforecastR::ciEnvelope(xmid, OCI[, 1], OCI[, 3], col = col1)
lines(xmid, OCI[, 2], col = "lightgrey", lwd = 4)

legend("topleft", legend = c("Model", "Data"), lwd = 10, col = c(col2, col1), lty = 1, cex = 1.7)
```

### Question 8

Question 8 [A]: Evaluate the model's ability to capture functional
responses to both Temperature and PAR.

> Looking at the point plot, observed NEE is a U-shaped curve
> with respect to temperature, whereas predicted NEE has a much
> higher, more constant envelope. This indicates that the model
> fails to capture the functional response to rising temperature.
>
> This is reflected clearly in the binned chart, where the model
> shows a positive bias on NEP, but around 10C the model and
> observations sharply diverge, with the model predicting increased
> NEP as temperature goes up, and observation showing reduced NEP
> as temperature goes up.


```{r}
plot(inputs$PAR[qaqc], O, pch = ".", ylab = "NEE")
points(inputs$PAR[qaqc], E, pch = ".", col = 2)

PAR = inputs$PAR[qaqc]
xd = seq(min(PAR),max(PAR),length=nbin)
xmid <- xd[-length(xd)] + diff(xd)
bin <- cut(PAR, xd)
Obar <- tapply(O, bin, mean, na.rm = TRUE)
Ose <- tapply(O, bin, std.error, na.rm = TRUE)
Ebar <- tapply(E, bin, mean, na.rm = TRUE)
Ese <- tapply(E, bin, std.error, na.rm = TRUE)
OCI <- -cbind(Obar - 1.96 * Ose, Obar, Obar + 1.96 * Ose)
ECI <- -cbind(Ebar - 1.96 * Ese, Ebar, Ebar + 1.96 * Ese)
rng <- range(rbind(OCI, ECI))

col2 <- ecoforecastR::col.alpha("darkgrey", 0.9)
col1 <- ecoforecastR::col.alpha("lightgrey", 0.6)

plot(xmid, Obar, ylim = rng, type = "n", xlab = "PAR", ylab = "NEP (umol/m2/s)", cex.lab = 1.3)
ecoforecastR::ciEnvelope(xmid, ECI[, 1], ECI[, 3], col = col2)
lines(xmid, ECI[, 2], col = "black", lwd = 4)
ecoforecastR::ciEnvelope(xmid, OCI[, 1], OCI[, 3], col = col1)
lines(xmid, OCI[, 2], col = "lightgrey", lwd = 4)

legend("topleft", legend = c("Model", "Data"), lwd = 10, col = c(col2, col1), lty = 1, cex = 1.7)
```

> In the point cloud, the range of observed NEE increases with PAR
> up to about 1000, but is evenly spread throughout its range for
> all PAR. The model shows a tightly constrained increase in NEE
> with increasing PAR. This shows that the model is not capturing
> an important functional response to PAR.
>
> The binned figure shows similar, with modeled NEP increasing almost
> linearly with PAR, whereas the observed response to increasing
> PAR levels out at about 1000, until 2200.

## Overall

Below is a final summary figure of the model's performance on a daily
timescale that combines many of the previous assessments.

```{r}
### other summary figures to go in multi-panel
par(mfrow = c(2, 2))

## Time-series visualization, daily means
DoY <- floor(L4$DoY - 0.02)
uDoY <- sort(unique(DoY))
ci.pf <- apply(apply(NEE.pf[, ], 2, tapply, DoY, mean), 1, mean)
NEE <- -L4$NEE_st_fMDS
NEEd <- tapply(NEE, DoY, mean)
plot(uDoY, ci.pf, xlab = "time", ylab = "NEE", type = "l", ylim = range(c(ci.pf, NEEd)), cex.lab = 1.3)
points(uDoY, NEEd, col = 2, pch = "+")
legend("topright", legend = c("Model", "Data"), lty = c(1, NA), pch = c(NA, "+"), col = 1:2, cex = 1.3)

## predicted vs observed
plot(NEEd, ci.pf, xlab = "Model", ylab = "Data", cex.lab = 1.3)
abline(0, 1, lty = 2, lwd = 4)
abline(lm(ci.pf ~ NEEd), col = 2, lwd = 3, lty = 3)
legend("topleft", legend = c("1:1", "Reg"), lty = 2:3, lwd = 4, col = 1:2, cex = 1.3)

## Functional response
plot(xmid, Obar, ylim = rng, type = "n", xlab = "Air Temperature (C)", ylab = "NEP (umol/m2/s)", cex.lab = 1.3)
ecoforecastR::ciEnvelope(xmid, ECI[, 1], ECI[, 3], col = col2)
lines(xmid, ECI[, 2], col = "black", lwd = 4)
ecoforecastR::ciEnvelope(xmid, OCI[, 1], OCI[, 3], col = col1)
lines(xmid, OCI[, 2], col = "lightgrey", lwd = 4)

legend("bottom", legend = c("Model", "Data"), lwd = 10, col = c(col2, col1), lty = 1, cex = 1.3)

### Classification tree
par(mar = c(0, 0, 0, 0))
rpe <- rpart(e2 ~ PAR + Tair, as.data.frame(x), method = "anova") ## sq error
plot(rpe, margin = 0.1)
text(rpe, cex = 1.5)
```

# Extra Material

## Comparison to flux "climatology"

In the section below we calculate the long-term average NEE for each 30
min period in the year, excluding the year we modeled (2005) as an
alternative model to judge our process model against. We then update our
summary statistics and predicted-observed plot

```{r}
## flux "climatology"
fluxfiles <- dir("data", pattern = "AMF")
fluxfiles <- fluxfiles[grep("txt", fluxfiles)]
fluxfiles <- fluxfiles[-grep("2005", fluxfiles)]
clim.NEE <- clim.doy <- NULL
for (f in fluxfiles) {
  ff <- read.csv(file.path("data", f), header = TRUE, na.strings = "-9999")
  ff[ff == -9999] <- NA
  clim.NEE <- c(clim.NEE, ff$NEE_st_fMDS)
  clim.doy <- c(clim.doy, ff$DoY)
}
NEE.clim <- tapply(clim.NEE, clim.doy, mean, na.rm = TRUE)[1:length(qaqc)]
C <- NEE.clim[qaqc]
NEE.clim.fit <- lm(O ~ C)
summary(NEE.clim.fit)
stats["RMSE", 3] <- sqrt(mean((C - O)^2))
stats["Bias", 3] <- mean(C - O)
stats["cor", 3] <- cor(C, O)
stats["slope", 3] <- coef(NEE.clim.fit)[2]
colnames(stats)[3] <- "clim"
knitr::kable(stats)
plot(C, O, pch = ".", xlab = "climatology", ylab = "observed", main = "NEE (umol/m2/sec)")
abline(0, 1, col = 2, lwd = 2)
abline(NEE.clim.fit, col = 3, lwd = 3, lty = 2)
legend("bottomright", legend = c("obs", "1:1", "reg"), col = 1:3, lwd = 3)

## example cycle
plot(L4$DoY, -L4$NEE_st_fMDS, xlim = c(200, 210), type = "l", lwd = 2, ylim = c(-10, 20), xlab = "Day of Year", ylab = "NEE")
lines(L4$DoY, -NEE.clim, col = 4, lwd = 2, lty = 2)
legend("topright", legend = c("Obs", "clim"), lty = 1:2, col = c(1, 4), lwd = 2)
```

### Question 9

Question 9 [B]: How does the process model perform relative to the
average flux data? Which statistics showed the largest differences
between the model and climatology?

## Time-scales

Many ecological processes operate at multiple time scales. For example,
carbon flux data responds to the diurnal cycle of light and temperature,
meso-scale variability due to weather fronts, seasonal variability, and
inter-annual variability driven by longer-term climate modes, as well as
disturbance and succession.

In the next section we look at the average diurnal cycle of the data and
models.

```{r}
## diurnal cycle
NEE.ens.diurnal <- tapply(E, L4$Hour[qaqc], mean)
NEE.pf.diurnal <- tapply(P, L4$Hour[qaqc], mean)
NEE.clim.diurnal <- tapply(C, L4$Hour[qaqc], mean)
NEE.obs.diurnal <- tapply(O, L4$Hour[qaqc], mean)
ylim <- range(c(NEE.ens.diurnal, NEE.pf.diurnal, NEE.obs.diurnal))
tod <- sort(unique(L4$Hour))
plot(tod, NEE.ens.diurnal, ylim = ylim, col = 2, xlab = "Time of Day", ylab = "NEE", main = "Diurnal Cycle", type = "l", lwd = 3)
lines(tod, NEE.pf.diurnal, col = 3, lwd = 3)
lines(tod, NEE.clim.diurnal, col = 4, lwd = 3)
lines(tod, NEE.obs.diurnal, lwd = 3)
legend("bottomright", legend = c("obs", "ens", "PF", "clim"), col = 1:4, pch = 20, cex = 0.75)
```


### Question 10

Question 10 [C]: What time of day has the largest uncertainty? What does
this suggest about what parameter(s) needs to be modified in the model,
in what direction, and by approximately how much? In providing this
answer, recall the structure of the model as well as the fact that the
particle filter has assimilated LAI so we can assume that that term is
unbiased for that case.
